Figure 4a and Supplementary Figure 5

### 2018-02-28 inst/tcga_survival_total/tcga_survival_total.Rmd
last updates before CDD submission

---
title: "Figure 4a-4b"
subtitle: "lncTAM34a survival analysis in TCGA cancers"
author: "Jimmy Van den Eynden and Jason T. Serviss"
date: "31/01/2017"
output:
  html_document:
    theme: flatly
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style> 
.caption { 
  color: Black; 
  font-family: "Times New Roman", Times, serif;
  font-size: 1.0em; 
}
</style>

```{r, echo = FALSE, message = FALSE}

packages <- c(
    "tidyverse",
    "printr",
    "ggthemes",
    "readr",
    "miR34AasRNAproject",
    "grid",
    "gtable"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

#library(grDevices)
```

@@

**_Please note that the following script has not yet been fully integrated into the package and, therefore, is included here to facilitate transparency although re-running the script in real-time is not yet supported. This will be resolved before publication._**

Data were derived from TCGA:

* **data/lncRNA_cancer_id.rds**: Matrix containing for each TCGA barcode information regarding cancer type.
* **data/lncRNA_clin.rds**: Matrix containing for each TCGA barcode information regarding vital state (first column) and date of last follow-up (second column). Data were downloaded from TCGA on 20171027.
* **data/lncRNA_TCGA_TP53_hasNonsynMut.rds**: Named vector containing for each TCGA barcode information regarding TP53 nonSynonymous mutation state: FALSE if no mutation or synonymous mutations, TRUE if mutated. Mutect2 v7 called data were downloaded from TCGA on 20171030.
* **data/lncRNA_expr_RP3.rds**: Named numeric vector containing for each TCGA barcode information regarding RP3-510D11.2 expression. For details on RNA-Seq processing, see *Ashouri A et al. Nat Commun. 2016;7*
* **data/lncRNA_cna_RP3.rds**: Named numeric vector containing for each TCGA barcode information regarding RP3-510D11.2 copy number alteration (gene-based segmentation values). For details on processing, see *Ashouri A et al. Nat Commun. 2016;7*
* **data/lncRNA_expr_miR34a.rds**: Named numeric vector containing for each TCGA barcode information regarding miR34a expression. For details on RNA-Seq processing, see *Ashouri A et al. Nat Commun. 2016;7*
* **data/lncRNA_TCGA_BRCA_PAM50.rds**: Named character vector containing for each TCGA barcode from the breast cancer dataset information regarding PAM50 classification: LumA, LumB, Basal or Her2. For details on PAM50 classification, see *Ashouri A et al. Nat Commun. 2016;7*

```{r, eval = FALSE}
TCGA_cancer_id<- readRDS("data/lncRNA_cancer_id.rds") # TCGA Subtype IDs
TCGA_clin_surv<- readRDS("data/lncRNA_clin.rds") # Survival data
TCGA_TP53_hasNonsynMut<- readRDS("data/lncRNA_TCGA_TP53_hasNonsynMut.rds") # Survival data
RP3_expr<- readRDS(file = "data/lncRNA_expr_RP3.rds")
RP3_cna<- readRDS(file = "data/lncRNA_cna_RP3.rds")
miR34a_expr<- readRDS(file = "data/lncRNA_expr_miR34a.rds")
TCGA_PAM50<- readRDS(file = "data/lncRNA_TCGA_BRCA_PAM50.rds")
```

## Functions

```{r, eval = FALSE}
source("functions/create_survival_plot.R") # Function used to create survival plots
source("functions/create_expression_groups.R") # Function used to divide samples based on expression percentile thresholds
source("functions/get_survival.R") # Function used to calculate survival probabilities
```

## Libraries
```{r, eval = FALSE}
  library(survival)
  library(KMsurv)
  library(svglite)
  library(WriteXLS)
```

# Survival analysis per cancer type

## Kaplan-Meier

```{r, message=F, warning=F, eval = FALSE}

    # Get cancers for which all expression data available: 17 in total
    cancers<- sort(names(table(TCGA_cancer_id[names(RP3_expr),1])))

    # Set expression th
    th1<- 0.1
    th2<- 1-th1

    p_vals<- NULL      
    # Create plots
  for(plot_format in 1:2){
    if(plot_format==1) pdf("figs/lncRNA_survival_KM.pdf",width = 17,height = 24)
    else svglite("figs/lncRNA_survival_KM.svg",width = 17,height = 24)
    par(mfrow=c(9,6))
    for(cancer in cancers){
      samples_cancer<- rownames(TCGA_cancer_id)[TCGA_cancer_id[,1]==cancer]

      # Mutation only for TP53
      TCGA_TP53_hasNonsynMut_temp<- TCGA_TP53_hasNonsynMut[names(TCGA_TP53_hasNonsynMut)%in%samples_cancer]
      names(TCGA_TP53_hasNonsynMut_temp)<- substr(names(TCGA_TP53_hasNonsynMut_temp),1,12)
      clin_class<- cbind(TCGA_clin_surv,TCGA_TP53_hasNonsynMut_temp[rownames(TCGA_clin_surv)])
      n1<- sum(clin_class[,3]==FALSE,na.rm=T)
      n2<- sum(clin_class[,3]==TRUE,na.rm=T)
      if(nrow(clin_class)==0|sum(!is.na(clin_class[,3]))==0) plot.new() # No use if no motation information
      else p1<- create_survival_plot(clinical_data = clin_class,classes_to_compare = c(FALSE,TRUE),classnames=c(paste("TP53 WT (n=",n1,")",sep=""),paste("TP53 MUT (n=",n2,")",sep="")),plot_title =  cancer,save_plot = TRUE,add_to_existing_plot = TRUE, plot_cols = c("black","black"), plot_lty = 1:2,return_p = TRUE)

    # Expr + Mutation
    for(i in 1:2){
      expr<- list(RP3_expr,miR34a_expr)[[i]]
      gene<- c("miR34a asRNA","miR34a")[i]
      # for(isMut in c(F,T)){
      isMut=FALSE
        samples_cancer_sel<- intersect(names(TCGA_TP53_hasNonsynMut)[TCGA_TP53_hasNonsynMut==isMut],samples_cancer)
        expr_temp<- expr[names(expr)%in%samples_cancer_sel]
        class<- create_expression_groups(expr_temp,gene,th=c(th1,th2))
        names(class)<- substr(names(class),1,12)
        class[class==3]<- 2
        clin_class<- cbind(TCGA_clin_surv[names(class),],class)
        if(sum(!is.na(clin_class[,3]))==0) next
        n1<- sum(clin_class[,3]==1,na.rm=T)
        n2<- sum(clin_class[,3]==2,na.rm=T)
        n3<- sum(clin_class[,3]==3,na.rm=T)
        # if(nrow(clin_class)==0|sum(!is.na(clin_class[,3]))==0|n1==0|n2==0|n3==0) plot.new()
        if(nrow(clin_class)==0|sum(!is.na(clin_class[,3]))==0|n1==0|n2==0) plot.new()
        else px<- create_survival_plot(clinical_data = clin_class,classes_to_compare = c(1,2),classnames=c(paste("Low expression (<P",100*th1," - n=",n1,")",sep=""),paste( "Normal expression (>P",100*th1," - n=",n2,")",sep="")),plot_title = paste(cancer,gene,sep=" "),save_plot = TRUE,add_to_existing_plot = TRUE,plot_cols = c("black","red"), plot_lty=rep(1+isMut,2),return_p = TRUE)
      # }
        if(i==1) p2<- px
        if(i==2) p3<- px
    }
      p_vals<- cbind(p_vals,c(p1,p2,p3))
    }
    dev.off()
  }

  p_vals<- p_vals[,1:17]
  colnames(p_vals)<- cancers
  rownames(p_vals)<- c("TP53","miR34a_asRNA","miR34a")
```

## 5y survival probabilities

```{r, message=F, warning=F, eval = FALSE}

# Create correlation matrix
common_samples<- intersect(names(TCGA_TP53_hasNonsynMut),names(RP3_expr))
correlation_matrix<- cbind(RP3=RP3_expr[common_samples],miR34a=miR34a_expr[common_samples],TP53=TCGA_TP53_hasNonsynMut[common_samples],cancer=TCGA_cancer_id[common_samples,1])

n_years<- 5

# Get surv probs
cancers<- sort(unique(correlation_matrix[,"cancer"]))
survProb_ls<- list()

# Independent
for(gene in c("RP3","miR34a","TP53")){
  cat(gene,"\n")
  surv_matrix<- matrix(NA,3,length(cancers),dimnames = list(1:3,cancers))
  for(cancer in cancers){
    correlation_matrix_temp<- correlation_matrix[correlation_matrix[,"cancer"]==cancer,]
    if(gene!="TP53") correlation_matrix_temp<- cbind(correlation_matrix_temp,class=create_expression_groups(t(correlation_matrix_temp),gene,th=c(0.1,0.9)))
    else correlation_matrix_temp<- cbind(correlation_matrix_temp,class=as.numeric(correlation_matrix_temp[,gene]=="TRUE")+1) 
    for(i in 1:2){
      if(i==1) samples_sel<- rownames(correlation_matrix_temp)[correlation_matrix_temp[,"class"]==i]
      else samples_sel<- rownames(correlation_matrix_temp)[correlation_matrix_temp[,"class"]%in%c(2,3)]
      if(length(samples_sel)<=1) next
      surv_data<- TCGA_clin_surv[rownames(TCGA_clin_surv)%in%substr(samples_sel,1,12),]
      survProb<- get_survival(surv_data,n_years)
      if(length(survProb)==0) next
      surv_matrix[i,cancer]<- survProb
    }
  }
  survProb_ls[gene]<- list(surv_matrix)
}

# TP53 status dependent
for(gene in c("RP3","miR34a")){
  cat(gene,"\n")
  for(mut in c(0,1)){
    surv_matrix<- matrix(NA,3,length(cancers),dimnames = list(1:3,cancers))
    for(cancer in cancers){
      correlation_matrix_temp<- correlation_matrix[correlation_matrix[,"cancer"]==cancer&as.numeric(correlation_matrix[,"TP53"]=="TRUE")==mut,,drop=F]
      if(nrow(correlation_matrix_temp)<=1) next
      correlation_matrix_temp<- cbind(correlation_matrix_temp,class=create_expression_groups(t(correlation_matrix_temp),gene,th=c(0.1,0.9)))
      for(i in 1:2){
        if(i==1) samples_sel<- rownames(correlation_matrix_temp)[correlation_matrix_temp[,"class"]==i]
        else samples_sel<- rownames(correlation_matrix_temp)[correlation_matrix_temp[,"class"]%in%c(2,3)]
        if(length(samples_sel)<=1) next
        surv_data<- TCGA_clin_surv[rownames(TCGA_clin_surv)%in%substr(samples_sel,1,12),]
        survProb<- get_survival(surv_data,n_years)
        if(length(survProb)==0) next
        surv_matrix[i,cancer]<- survProb
      }
    }
    survProb_ls[paste(gene,mut,sep="_")]<- list(surv_matrix)
  }
}

# Remove NA!
survProb_ls[["RP3"]]<- survProb_ls[["RP3"]][,!is.na(survProb_ls[["RP3"]][1,])&!is.na(survProb_ls[["RP3"]][2,])]
survProb_ls[["miR34a"]]<- survProb_ls[["miR34a"]][,!is.na(survProb_ls[["miR34a"]][1,])&!is.na(survProb_ls[["miR34a"]][2,])]
survProb_ls[["TP53"]]<- survProb_ls[["TP53"]][,!is.na(survProb_ls[["TP53"]][1,])&!is.na(survProb_ls[["TP53"]][2,])]
survProb_ls[["RP3_0"]]<- survProb_ls[["RP3_0"]][,!is.na(survProb_ls[["RP3_0"]][1,])&!is.na(survProb_ls[["RP3_0"]][2,])]
survProb_ls[["RP3_1"]]<- survProb_ls[["RP3_1"]][,!is.na(survProb_ls[["RP3_1"]][1,])&!is.na(survProb_ls[["RP3_1"]][2,])]
survProb_ls[["miR34a_0"]]<- survProb_ls[["miR34a_0"]][,!is.na(survProb_ls[["miR34a_0"]][1,])&!is.na(survProb_ls[["miR34a_0"]][2,])]
survProb_ls[["miR34a_1"]]<- survProb_ls[["miR34a_1"]][,!is.na(survProb_ls[["miR34a_1"]][1,])&!is.na(survProb_ls[["miR34a_1"]][2,])]
# 
# for(plot_format in 1:2){
#   if(plot_format==1) pdf("figs/lncRNA_5ySurv_expr.pdf",width = 14, height = 14)
#   else svglite("figs/lncRNA_5ySurv_expr.svg",width = 14, height = 14)
# 
#   par(mfrow=c(2,1))
# 
#   boxplot(list(
#     survProb_ls[["RP3"]][1,],survProb_ls[["RP3"]][2,],NA,
#     survProb_ls[["miR34a"]][1,],survProb_ls[["miR34a"]][2,],NA,
#     survProb_ls[["TP53"]][2,],survProb_ls[["TP53"]][1,]),
#     outline=F,ylab=paste(n_years,"y survival probability",sep=""),frame.plot=F,axes=F,col=c("lightgrey",NA,NA,"lightgrey",NA,NA,"darkgrey",NA)
#   )
#   axis(2)
#   text(1.5,1,paste("P=",signif(wilcox.test(survProb_ls[["RP3"]][1,],survProb_ls[["RP3"]][2,],paired=T)$p.value,3),sep=""))
#   text(4.5,1,paste("P=",signif(wilcox.test(survProb_ls[["miR34a"]][1,],survProb_ls[["miR34a"]][2,],paired=T)$p.value,3),sep=""))
#   text(7.5,1,paste("P=",signif(wilcox.test(survProb_ls[["TP53"]][1,],survProb_ls[["TP53"]][2,],paired=T)$p.value,3),sep=""))
#   text(1.5,1.05,"miR34a asRNA",xpd=NA)
#   text(4.5,1.05,"miR34a",xpd=NA)
#   text(7.5,1.05,"TP53",xpd=NA)
#   legend("bottom",horiz = T,legend = c("Control","Low expression (<P10)","Mutated"),fill = c("white","lightgrey","darkgrey"),bty="n")
#   
#   boxplot(list(
#     survProb_ls[["RP3_0"]][1,],survProb_ls[["RP3_0"]][2,],NA,
#     survProb_ls[["RP3_1"]][1,],survProb_ls[["RP3_1"]][2,],NA,
#     survProb_ls[["miR34a_0"]][1,],survProb_ls[["miR34a_0"]][2,],NA,
#     survProb_ls[["miR34a_1"]][1,],survProb_ls[["miR34a_1"]][2,]),
#     outline=F,ylab=paste(n_years,"y survival probability",sep=""),frame.plot=F,axes=F,col=c("lightgrey",NA,NA,"lightgrey",NA,NA,"lightgrey",NA,NA,"lightgrey",NA)
#   )
#   axis(2)
#   text(1.5,1,signif(wilcox.test(survProb_ls[["RP3_0"]][1,],survProb_ls[["RP3_0"]][2,],paired=T)$p.value,3))
#   text(4.5,1,signif(wilcox.test(survProb_ls[["RP3_1"]][1,],survProb_ls[["RP3_1"]][2,],paired=T)$p.value,3))
#   text(7.5,1,signif(wilcox.test(survProb_ls[["miR34a_0"]][1,],survProb_ls[["miR34a_0"]][2,],paired=T)$p.value,3))
#   text(10.5,1,signif(wilcox.test(survProb_ls[["miR34a_1"]][1,],survProb_ls[["miR34a_1"]][2,],paired=T)$p.value,3))
#   text(3,1.1,"miR34a asRNA",xpd=NA)
#   text(1.5,1.05,"TP53 wt",xpd=NA)
#   text(4.5,1.05,"TP53 mut",xpd=NA)
#   text(9,1.1,"miR34a",xpd=NA)
#   text(7.5,1.05,"TP53 wt",xpd=NA)
#   text(10.5,1.05,"TP53 mut",xpd=NA)
#   legend("bottom",horiz = T,legend = c("Control","Low expression (<P10)"),fill = c("white","lightgrey"),bty="n")
#   
#   dev.off()
# }

# Correlation between survival
 cancer_colors<- c(
  rgb(0.498,0.498,0.498),
  rgb(0.7804,0.7804,0.7804),
  rgb(0.1216,0.4667,0.7059),
  rgb(0.1216,0.4667,0.7059),
  rgb(0.6824,0.7804,0.9098),
  rgb(0.0902,0.7451,0.8118),
  rgb(0.6196,0.8941,0.898),
  rgb(1,0.498,0.0549),
  rgb(1,0.7333,0.4706),
  rgb(0.8392,0.1529,0.1569),
  rgb(1,0.5961,0.5882),
  rgb(0.5804,0.4039,0.7412),
  rgb(0.7725,0.6118,0.7961),
  rgb(0.8902,0.4667,0.7608),
  rgb(0.9686,0.7137,0.8235),
  rgb(0.549,0.3373,0.2941),
  rgb(0.7686,0.6118,0.5804),
  rgb(0.1725,0.6275,0.1725),
  rgb(0.5961,0.8745,0.5412),
  rgb(0.7373,0.7412,0.1333),
  rgb(0.8588,0.8824,0.5529),
  rgb(0.9294,0.8,0.0588),
  rgb(1,0.9255,0)
)
  names(cancer_colors)<- c('ACC','BLCA','BRCA','BRCA Basal','BRCA Her2','BRCA LumA','BRCA LumB','CESC','GBM','HNSC','KICH','KIRC','KIRP','LGG','LIHC','LUAD','LUSC','OV','PRAD','SKCM','STAD','THCA','UCS')

for(plot_format in 1:2){
  if(plot_format==1) pdf("figs/lncRNA_5ySurv_expr_corr.pdf",width = 14, height = 7)
  else svglite("figs/lncRNA_5ySurv_expr_corr.svg",width = 21, height = 7)

  par(mfrow=c(1,3))

  # Focus on WT group only!
  survProb_rel_RP3<- survProb_ls[["RP3_0"]][1,]-survProb_ls[["RP3_0"]][2,]
  survProb_rel_miR34a<- survProb_ls[["miR34a_0"]][1,]-survProb_ls[["miR34a_0"]][2,]
  survProb_rel_TP53<- survProb_ls[["TP53"]][2,]-survProb_ls[["TP53"]][1,]

  for(i in 1:3){
    x_param_name<- c("TP53","miR34a_asRNA","miR34a")[i]
    x_param<- list(survProb_rel_TP53,survProb_rel_RP3,survProb_rel_miR34a)[[i]]
    y_param_name<- c("miR34a_asRNA","miR34a","TP53")[i]
    y_param<- list(survProb_rel_RP3,survProb_rel_miR34a,survProb_rel_TP53)[[i]]
    common_cancers<- intersect(names(x_param),names(y_param))
    x_param<- x_param[common_cancers]
    y_param<- y_param[common_cancers]
    plot(x_param,y_param,frame.plot=F,xlim=c(-1,+1),ylim=c(-1,+1),xlab=x_param_name,ylab=y_param_name)  
    boxplot(y_param,at = 0.75,add=T,frame.plot=F,outline=F,boxwex=0.2)
    boxplot(x_param,at = 0.75,add=T,frame.plot=F,outline=F,horizontal = T,boxwex=0.2)
    p_temp<- signif(cor.test(x_param,y_param,method = "spearman")$p.value,3)
    r_temp<- signif(cor.test(x_param,y_param,method = "spearman")$estimate,3)
    legend("topleft",legend = c(paste("P=",p_temp,sep=""),paste("r=",r_temp,sep="")),bty="n")
    for(cancer in common_cancers){
      points(x_param[cancer],y_param[cancer],col=cancer_colors[cancer],pch=16,cex=2)
    }
    # abline(h=0,v=0,lty=2)
  }
  dev.off()
}

  # Survival effects
  median(survProb_rel_RP3) # -0.096
  median(survProb_rel_miR34a) # -0.068
  median(survProb_rel_TP53) # -0.064
  wilcox.test(survProb_rel_RP3) # p=0.083
  wilcox.test(survProb_rel_miR34a) # p=0.17
  wilcox.test(survProb_rel_TP53) # p=0.04

@@

>>>>

---
title: "Figure 4a and Supplementary Figure 5"
subtitle: "Survival curves"
author: "Jimmy Van den Eynden and Jason T. Serviss"
date: "31/01/2017"
output:
  html_document:
    theme: flatly
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

<style>
.col2 { 
    columns: 2 200px; /* number of columns and width in pixels*/ 
    -webkit-columns: 2 200px; /* chrome, safari */ 
    -moz-columns: 2 200px; /* firefox */ 
}
</style>

```{r, echo = FALSE, message = FALSE}
packages <- c(
  "tidyverse",
  "printr",
  "ggthemes",
  "readr",
  "miR34AasRNAproject"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

@@

#### _Introduction_
Kaplan-Meier survival curves comparing the effects of TP53-mutated samples,
low lncTAM34a expression and low miR34a expression to control samples in
17 cancers from TCGA. Panels showing survival curves based on gene expression
include only TP53 wild type patients where RNAseq data exists.

#### _Methods_
RNAseq data was downloaded from TCGA and processed as described previously. 
Briefly, RNAseq data were aligned to the human hg19 assembly and quantified 
using GENCODE (v19) annotated HTSeq-counts and FPKM normalizations. Expression 
data from miR34a and lncTAM34a (identified as RP3-510D11.2) were used for 
further analysis.

Survival analysis was performed on TCGA vital state and follow-up data,
downloaded from GDC on 27/10/2017 using the R survival package.

<br></br>

#### _Results_
```{r}

url <- fileMap(type = "rds")["Figure 4b"][[1]]
data <- read_rds(gzcon(url(file.path(dataUrl, url))))

@@

```{r}
#TP53 status independant
#Set up all combinations of variables to calcualte survival probability for
independant <- expand.grid(
  gene = c("lncTAM34a", "miR34a", "TP53"),
  cancer = sort(unique(data$Cancer))
) %>% 
  as_tibble() %>%
  mutate_if(is.factor, as.character) %>%
  group_by(gene, cancer) %>%
  #add data of the corresponding cancer and, for gene expression, filter WT TP53
  mutate(data = map2(cancer, gene, function(c, g, cm = data) {
    if(g == "TP53") {
      filter(cm, Cancer == c)
    } else {
      filter(cm, Cancer == c & TP53 == FALSE)
    }
  })) %>%
  #select the gene to be examined
  mutate(data = pmap(list(gene, data), function(g, d) {
      select(d, g, vitalStatus, FU)
  })) %>%
  #remove samples where gene expression/mutation status is na
  mutate(data = map(data, function(x) filter(x, is.na(x[[1]]) == FALSE))) %>%
  #remove cancers where comparisons without data exist
  mutate(bool = map_lgl(data, function(x) nrow(x) == 0)) %>%
  group_by(cancer) %>%
  filter(!any(bool)) %>%
  #calculate the percentile for the gene's expression for each sample
  mutate(data = map2(gene, data, function(g, d) {
    if(any(grepl("TP53", colnames(d)))) {
      mutate(d, percentile = NA)
    } else {
      mutate(d, percentile = ntile(x = d[[1]], n = 10))
    }
  })) %>%
  #classify 10th percentile and under as "low" and the rest as "others"
  #classify TP53 mutated as low and WT as others
  mutate(data = map(data, function(d) {
    if(any(grepl("TP53", colnames(d)))) {
      mutate(d, class = if_else(TP53, "low", "others"))
    } else {
     mutate(d, class = if_else(percentile == 1, "low", "others")) 
    }
  })) %>%
  #calculate survival probability
  mutate(surv.obj = map(data, ~get_survival(.x))) %>%
  mutate(surv.fit = map2(data, surv.obj, function(x, y) survival_fit(y, x$class))) %>%
  mutate(p.value = map2_dbl(data, surv.obj, function(x, y) survival_p(y, x$class))) %>%
  ungroup()
```

<div class="col2">
```{r}
trash <- pmap(
  list(independant$data, independant$surv.fit, independant$gene, independant$cancer, independant$p.value), 
  function(v, w, x, y, z) {
    plotKM(v, w, x, y, z)
  }
)
```
</div>
